---
title: "Joint models anushka"
author: "Anushka Muley"
date: "2025-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
suppressPackageStartupMessages({library(ggplot2)
library(MASS)
library(gtsummary)
library(dplyr)
library(survival)
library(survminer)
library(knitr)
library(sjPlot)
library(JMbayes2)
library(splines)})
```


## Load package


```{r}
# PBC data (JMbayes2)
data("pbc2", package = "JMbayes2")
data("pbc2.id", package = "JMbayes2")

pbc <- pbc2 %>%
  mutate(id = as.integer(id),
         sex = factor(sex)) %>%
  arrange(id, year)

# survival frame with explicit {0,1} event indicator
pbc_surv <- pbc2.id %>%
  mutate(
    id = as.integer(id),
    event = as.integer(status2 != 0),
    sex = factor(sex) 
  )
```
#Descriptive Statistics

```{r}
library(dplyr)

pbc %>%
  summarise(
    n = n(),
    mean_age = mean(age, na.rm = TRUE),
    sd_age = sd(age, na.rm = TRUE),
    median_bili = median(serBilir, na.rm = TRUE),
    min_bili = min(serBilir, na.rm = TRUE),
    max_bili = max(serBilir, na.rm = TRUE)
  )

```
# Summary Table

```{r}
library(gtsummary)

pbc %>%
  select(age, status, sex, serBilir, albumin, drug, year, ascites, edema) %>%
  tbl_summary(
    missing = "always"
  ) %>%
  modify_header(label ~ "**Variable**") %>%
  bold_labels()

```



#Longitudinal Spagetti

```{r}
p_pbc_long <- ggplot(pbc, aes(x = year, y = log(serBilir), group = id)) +
  geom_line(alpha = 0.2) +
  stat_smooth(aes(group = 1), method = "loess", se = TRUE, linewidth = 1) +
  labs(title = "PBC: log(serum bilirubin) trajectories",
       x = "Years since baseline", y = "log(serum bilirubin)")
p_pbc_long

```



# Plot for visualisation

```{r}
ggplot(pbc, aes(year, log(serBilir), group = id)) +
  geom_line(alpha = 0.15) +
  facet_wrap(~ sex) +
  labs(title = "Log(bilirubin) trajectories by sex",
       x = "Years", y = "log(bilirubin)")

```
```{r}
ggplot(pbc, aes(year, log(serBilir), group = id)) +
  geom_line(alpha = 0.15) +
  facet_wrap(~ drug) +
  labs(title = "Log(bilirubin) trajectories by drug",
       x = "Years", y = "log(bilirubin)")

```
Kaplan–Meier by baseline bilirubin (tertiles)
```{r, fig.width=8,fig.height=8}
pbc_surv <- pbc2.id %>%
  mutate(id = as.integer(id),
         event = as.integer(status2 != 0))

pbc_base <- pbc %>%
  group_by(id) %>%
  arrange(year, .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(bili_grp = ntile(serBilir, 3),
         bili_grp = factor(bili_grp, labels = c("Low","Medium","High")))

# Now join works
km_df <- pbc_surv %>%
  left_join(pbc_base %>% select(id, bili_grp), by = "id") %>%
  filter(!is.na(bili_grp))

fit_km_pbc <- survfit(Surv(years, event) ~ bili_grp, data = km_df)

ggsurvplot(
  fit_km_pbc, data = km_df, risk.table = TRUE,
  title = "PBC: Survival by baseline serum bilirubin tertiles",
  legend.title = "Baseline ser. bilirubin"
)

```
# Longitudinal Model

```{r}
# LME (log serBilir ~ nonlinear time + covariates)
model1 <- nlme::lme(
  log(serBilir) ~ year + sex + age,
  random = ~ year | id,
  data = pbc, na.action = na.omit,
  control = nlme::lmeControl(opt = "optim")
)
class(model1)
summary(model1)
is.null(model1)
inherits(model1, "try-error")

```
# Survival Model

```{r}
# Cox (time to event)
cox_pbc <- coxph(Surv(years, event) ~ sex + age, data = pbc_surv, x = TRUE, y = TRUE)
summary(cox_pbc)
#convert categorical variable into factor
pbc_surv <- pbc_surv %>%
  mutate(sex = factor(sex, labels = c("Female", "Male")))
```

The bilirubin increases non linearly over time.
The females have stighly lower bilirubin than males. Since p value is large we fail reject the null hypothesis that there is a difference in bilirubin levels of males and females.


# SJ Plot
```{r}
# Longitudinal Model 1: linear time
model1 <- nlme::lme(
  log(serBilir) ~ year + sex + age,
  random = ~ year | id,
  data = pbc,
  na.action = na.omit,
  control = nlme::lmeControl(opt = "optim")
)

# Longitudinal Model 2: spline nonlinear time 
model2 <- nlme::lme(
  log(serBilir) ~ ns(year, 2) + sex + age,
  random = ~ ns(year, 2) | id,
  data = pbc,
  na.action = na.omit,
  control = nlme::lmeControl(opt = "optim")
)

# Longitudinal Model 3: spline nonlinear time 
model3 <- nlme::lme(
  log(serBilir) ~ ns(year, 3) + sex + age,
  random = ~ ns(year, 3) | id,
  data = pbc,
  na.action = na.omit,
  control = nlme::lmeControl(opt = "optim")
)


tab_model(model1, model2, model3,
          dv.labels = c("Model 1", "Model 2", "Model 3"),
          show.aic = TRUE,
          show.se = TRUE,
          show.ci = FALSE)

```

#Joint Model

```{r}

```

